An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-05-13 manuscript/main.tex
latest draft of manuscript

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit as intended certain associations between item attributes and functioning (e.g. item complexity predicts item difficulty). However, we also find that item clones are not psychometrically equivalent and therefore not exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials and results made available here will encourage researchers to use the MaRs-IB in their studies.}

@@

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 unique matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing abstract shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to various arbitrary abstract rules (e.g. color change, position change). To solve a puzzle, participants deduce which of four response options correctly completes the matrix based on these rules. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of elements across cells (Figure \ref{fig:fig00}a/b). To increase the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. A single item's clones share identical structural characteristics (i.e. number of elements, number of rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses several of the issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB. A sample of 659 adult, adolescent, and child participants completed an experiment in which they had eight minutes to complete as many MaRs-IB items as possible (items were presented in the same order to all participants). The authors found that performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating good convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} publicly released a summary of the difficulty (i.e. proportion correct responses) of each item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations in the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics, and therefore impede the design of new matrix reasoning tasks using the MaRs-IB. The first is that, given the fixed-order task design,  there was relatively few responses collected for the majority of the items in the MaRs-IB. Indeed, only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of how much information an item provides about ability. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test (i.e. an assemblage of items) in measuring performance across a range of ability levels. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given experimenter-defined constraints (e.g. test duration, test difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory} we investigate how item complexity, i.e. the number of elements and number of rules, explain item functioning (i.e. difficulty, discrimination). We modeled these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Thus this explanatory analysis provides another means by which to validate the success of the MaRs-IB design; that is, to determine whether the MaRs-IB exhibits similar relationships between item characteristics and functioning as observed in other established matrix reasoning measures. As a secondary benefit, explanatory IRT models yield more accurate estimates of item parameters \citep{neuhaus2006separating}, which in turn helps to ensure that new MaRs-IB tests designed using these item parameter estimates function as expected.

A contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or whether clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and discrimination can be generated simply by swapping out one set of clones for another (e.g. exchanging clones from shape set 1 for shape set 2). To test for exchangeability in the MaRs-IB, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item functioning across clones by distractor type and shape set, which should be negligible if clones are to be treated as exchangeable.

@@

All data, code, and model outputs (including the item parameter estimates) are publicly available at: \url{https://github.com/dawlab/mars-irt}.

@@

The presentation of each item was preceded by a fixation cross, which lasted for 1200 ms. Upon presentation of the item, participants were given up to 30 s to solve the puzzle. After 25 s elapsed, a clock appeared to count down the remaining 5 s. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were instructed to respond carefully, but to guess if they could not solve the puzzle before the timer ran out. The format of the experiment and instructions were adapted from code publicly released as part of the original study (\url{https://app.gorilla.sc/openmaterials/36164}). The task was programmed in jsPsych \citep{de2015jspsych} and distributed using custom web-application software (available at \url{https://github.com/nivlab/nivturk}).

In order to ensure sufficient sampling of every item (and their clones) in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \cite{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As a consequence, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each item by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or or paired difference, PD; see \cite{chierchia2019matrix} for details).

@@

In our analyses, we considered two item template (level 1) attributes, element number and rule number, which have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes of these shapes across cells. In the MaRs-IB, there are four principal rules: size change, color change, position change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Importantly, element and rule number are mostly uncorrelated across item templates ($\rho$ = 0.176). The feature and rule number for each item was determined via manual annotation.

@@

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute. As described above, every item template in the MaRs-IB has three unique versions that differ in the geometric shapes populating its cells. There are 45 unique geometric shapes in total, organized into nine stimulus sets, where each item clone draws a subset of shapes from one of the nine sets (see Figure S2 for examples). As such, the groups in tripartite definition of shape set do not constitute meaningful categories and we therefore do not incorporate them as fixed effects. Furthermore, item clones drawing from the same shape set are not always populated by the same geometric shapes, making complicated the possibility of modeling the nine stimulus sets instead. The most rigorous way of modeling the presence of a given geometric shape on item functioning would be to include each as a binary attribute in estimating item difficulty and discrimination. In order to keep the models simple, we elected to leave shape set unmodeled and accounted for by the residual variability terms instead.

@@

The AMIS modeling framework provides a natural means of accomplishing the three goals of the calibration study. Through any of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses about the impact of these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can evaluate whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, an association between distractor type and item functioning would entail that clones are not psychometrically equivalent. Moreover, if the level-2 residual variance in item difficulty or discrimination were non-negligible, that would similarly entail that item clones are not psychometrically equivalent due to one or more unmodeled features (e.g. low-level perceptual properties).

@@

The results of the linear mixed-effects model of response times are summarized in Table \ref{table:1}. On average, participants spent 15.9 s (sd = 7.2 s) per item. There were significant main effects of accuracy ($\beta = 0.067$, $t = 9.84$, $p < 0.001$), rest score ($\beta = 0.125$, $t = 16.95$, $p < 0.001$), and item difficulty ($\beta = 0.137, t = 44.33, p < 0.001$). In other words, response times were slower on average for correct responses, better- performing participants, and more difficult items. The positive association between response time and accuracy is consistent with speed-accuracy trade-offs; that is, slower responses are associated with more careful work. This interpretation is corroborated by the main effect of rest score, such that more accurate participants were slower overall.

@@

In sum, we found evidence of speed-accuracy trade-offs in performance in this sample. Accurate responses and more accurate participants were both slower overall. Notably, better-performing participants demonstrated adjustments to their response times as function of item difficulty; whereas the lowest scoring participants maintained the same work rate regardless of item difficulty, the highest scoring participant exhibited the largest slowing in responses as items became more challenging. We return to this pattern of results in the General Discussion.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. By comparison, Model 5 demonstrated smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). This result indicates the presence of non-negligible residual variance in item discrimination across item templates but not clones. It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 6 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that Model 6 fit clearly better than Model 4 ($\Delta$ LOCO = 85.91, se = 6.33, supporting the inclusion of varying item discrimination parameters). We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

@@

Contrary to the assumption that item clones are exchangeable, we found an association between distractor type and item difficulty ($\delta_{\beta_3}$ = 1.105, 95\% HDI = 0.940 -- 1.269). Items with MD distractors were associated with an estimated 19.9\% reduction in accuracy compared to their equivalent items with PD distractors. Together, element rule, rule number, and distractor type explained slightly over half (52.0\%) of the variance in item difficulty. Regardless, a model that included a clone-level random effects term for item difficulty was preferred to models without. Therefore, the results indicate that item clones are not equally difficult and are thus not exchangeable.

The estimated item difficulty per clone, organized by distractor type and shape set, is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast, the pattern of item difficulties across clones by shape set is more complicated (Figure \ref{fig:fig02}b). After controlling for distractor type and mean response time, the magnitude of the clone-level residual variance is $\epsilon_{\beta k}$ = 0.620 (95\% HDI = 0.538 -- 0.701). Across the 128 unique sets of three clones, organized by item and distractor type, that level of residual variance translates into an average difference in difficulty of 0.752 (95\% HDI = 0.630 -- 0.877) between the easiest and hardest item clones of each triad, an effect that is two-thirds as large as the effect of distractor type. Furthermore, the shape set membership of the most difficult clone varies considerably across the 128 sets of clones (shape set 1 = 39; shape set 2 = 32; shape set 3 = 57). Thus, the sources of residual variability in item difficulty across clones is complex and likely reflect the idiosyncratic contributions of their low-level perceptual features.

Turning to item discrimination, we did not find a credible association between discrimination and distractor type ($\delta_{\alpha_3}$ = -0.029, 95\% HDI = -0.065 -- 0.005). Moreover, a model that included a clone-level random effects term for item discrimination was not preferred to a simpler model without. As such, there is not sufficient evidence to refute that the assumption that item clones are similarly discriminating. We note this result should be interpreted with caution given the relatively weak predictive improvement of the best-fitting model compared to a model with a clone-level random effects term.

Finally, we inspected the relationship between mean response time and item functioning. As expected, there was a positive association between mean response time and item difficulty ($\delta_{\beta_4}$ = 0.541, 95\% HDI = 0.414 -- 0.671). That is, response times were slower on average for more difficult items. There were not, however, a credible associations between item discrimination and average response time ($\delta_{\alpha_4}$ = -0.010, 95\% HDI = -0.030 -- 0.009).

@@

In the best-fitting model, there were several credible associations between person-level attributes and ability. There was a negative association between age and ability ($\rho_1$ = -0.299, 95\% HDI = -0.353 -- -0.246). There was also a small association between gender and ability ($\rho_2$ = 0.128, 95\% HDI = 0.078 -- 0.182), such that women performed marginally better than men. Average response time was positively associated with ability ($\rho_3$ = 0.427, 95\% HDI = 0.374 -- 0.478), as was $\Delta$ response time ($\rho_4$ = 0.368, 95\% HDI = 0.313 -- 0.424). In other words, participants with higher levels of ability spent on average more time on each item and deliberated for longer on more difficult items. Together, these four person attributes explained 50.3\% of the variance in ability. These results are consistent with a speed-accuracy trade-off --- higher levels of ability in this sample reflect in part a tendency to slow down and respond carefully overall and increasingly for more challenging items.

@@

Finally, we validated the fit of the best-fitting model using two posterior predictive model checking measures. The PPP-value for chi-square discrepancy statistic did not exceed the critical threshold ($\chi^2_{NC}$ = 15.904, PPP-value = 0.297), indicating that the model was sufficiently able to reproduce the distribution of observed scores. The PPP-value for the SGDDM statistic did exceed the critical threshold (PPP-value = 0.019). However, the mean of the realized SGDDM values (SGDDM = 0.0180) were only marginally larger than the mean of the posterior predicted SGDDM values (SGDDM = 0.0175). That is, the residual inter-item correlations in the data unexplained by the best-fitting model were small on average and only slightly larger than what we would expect to observe by chance. As such, there is little evidence for local dependence in the data. Overall then, we can conclude the best-fitting model provides an adequate fit to the data.

@@

In this first study, we recruited a large sample of adults to complete items from the MaRs-IB, making sure that each item (and its clones) were completed by a sufficient number of participants. Using multilevel item structure models, we then estimated the psychometric properties (i.e. difficulty and discrimination) of each item. Corroborating the results of an initial investigation of the MaRs-IB \citep{chierchia2019matrix}, we found that the items in the MaRs-IB vary greatly in their difficulty -- a prerequisite for measuring nonverbal reasoning across the ability spectrum. We also found that the items in the MaRs-IB exhibited medium-to-large levels of discrimination, similar to what has been found in investigations of other of matrix reasoning tasks \citep{chiesi2012using, chiesi2012item, van2013shortened}.

We also investigated how the structural characteristics of items impact their functioning. We found that element number and rule number were both positively associated with item difficulty. These two attributes exerted approximately equal influence on difficulty (i.e. a one-unit change in either was independently associated with a roughly 10\% reduction in performance). This finding is interesting in light of previous investigations of nonverbal reasoning tasks, which have found either the number of elements \citep{bethell1984adaptive} or number of rules \citep{mulholland1980components} to be more determinant of item difficulty. One possible reason for this discrepancy is that these two attributes are largely uncorrelated ($\rho$ = 0.176) across items in the MaRs-IB. Together element number and rule number explained 67.6\% of the variance in item difficulty across item templates (and 38.6\% of the variance across item clones), which is in line with previous investigations of matrix reasoning tasks \citep{carpenter1990one, matzen1994error}. These findings further validate the design of the MaRs-IB insofar that item complexity strongly determines item difficulty.

>>>>

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit clear associations between item complexity and difficulty. However, we also find that item clones are not always psychometrically equivalent and cannot assumed to be exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials and results made available here will encourage researchers to use the MaRs-IB in their studies.}

@@

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing geometric shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to a number of abstract rules. Based on these results, participants must deduce which of four response options correctly completes the matrix. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of these elements across cells (Figure \ref{fig:fig00}a/b). To increase the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. A single item's clones share identical structural characteristics (i.e. number of elements, number of rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses several of the issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB in which a sample of 659 adult, adolescent, and child participants had eight minutes to complete as many MaRs-IB items as possible (items were presented in the same order to all participants). The authors found that performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating satisfactory convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} made publicly available a summary of the difficulty (i.e. proportion correct responses) of each item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations in the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics, and therefore impede the design of new matrix reasoning tasks using the MaRs-IB. The first is that, given the fixed-order task design,  there was relatively few responses collected for the majority of the items. Indeed only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of how much information an item provides about ability. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test in measuring performance across a range of ability levels. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given experimenter-defined constraints (e.g. test duration and/or difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory}, we investigate how item complexity, i.e. the number of elements and number of rules, determine item functioning (i.e. difficulty, discrimination). We modeled these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Thus this explanatory analysis provides another means by which to validate the success of the MaRs-IB design; that is, to confirm whether the MaRs-IB exhibits similar relationships between item properties and functioning as observed in other established matrix reasoning measures. As a secondary benefit, explanatory IRT models yield more accurate estimates of item parameters \citep{neuhaus2006separating}, which in turn helps to ensure that new MaRs-IB tests, designed using the item parameter estimates from these IRT models, function as expected.

A third contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or if clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and discrimination can be generated simply by swapping out one set of item clones for another (e.g. exchanging clones that differ only in shape set). To test for exchangeability in the MaRs-IB, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item functioning across clones by distractor type and shape set, which must be negligible if clones are to be treated as exchangeable.

@@

All data, code, and model outputs (including the item parameter estimates) are publicly available at: \url{https://github.com/ndawlab/mars-irt}.

@@

The presentation of each item was preceded by a fixation cross, which lasted for 1200 ms. Upon presentation of the item, participants were given up to 30 s to solve the puzzle. After 25 s elapsed, a clock appeared to count down the remaining 5 s. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were instructed to respond carefully, but to guess if they could not solve the puzzle before the timer ran out. The format of the experiment and instructions were adapted from code publicly released as part of the original study (\url{https://app.gorilla.sc/openmaterials/36164}). The task was programmed in jsPsych \citep{de2015jspsych} and distributed to participants via the web using custom software (available at \url{https://github.com/nivlab/nivturk}).

In order to ensure sufficient sampling of every item template and clone in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \citealt{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As a consequence, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each item by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or or paired difference, PD; see \citealt{chierchia2019matrix} for details).

@@

In our analyses we considered two item template (level 1) attributes, element number and rule number, which have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes of these shapes across cells. In the MaRs-IB there are four principal rules: size change, color change, position change, and shape change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Element and rule number are uncorrelated across item templates ($\rho$ = 0.176), and were determined for each item via manual annotation.

@@

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute for the following reasons. As described above, every item template in the MaRs-IB has three unique versions that differ only in the geometric shapes populating its cells. There are 45 unique geometric shapes in total, organized into nine stimulus sets, where each item clone draws a subset of shapes from one of the nine sets (see Figure S2 for examples). As such, shape set --- indicating whether a clone is the 1st, 2nd, or 3rd version of an item template --- is not a meaningful label and we therefore do not incorporate it into the models. Furthermore, item clones drawing from the same shape set are not always populated by the same geometric shapes, making complicated the possibility of modeling the nine stimulus sets instead. The most rigorous way of modeling the presence or absence of a given geometric shape on item functioning would be to include each as a binary attribute predicting item difficulty and discrimination. In order to keep the models simple, we elected to leave shape set unmodeled and accounted for by the residual variability terms instead.

@@

The AMIS modeling framework provides a natural means of accomplishing the three objectives of the calibration study. Through all of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses about the impact of these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can determine whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, an association between distractor type and either item parameter would entail that clones are not psychometrically equivalent. Moreover, if either level-2 residual variance term were non-negligible, that would similarly entail that not all item clones are equivalent due to one or more unmodeled features (e.g. low-level perceptual properties).

@@

The results of the linear mixed-effects model of response times are summarized in Table \ref{table:1}. On average, participants spent 15.9 s (sd = 7.2 s) per item. There were significant main effects of accuracy ($\beta = 0.067$, $t = 9.84$, $p < 0.001$), rest score ($\beta = 0.125$, $t = 16.95$, $p < 0.001$), and item difficulty ($\beta = 0.137, t = 44.33, p < 0.001$). In other words, response times were slower on average for correct responses, better- performing participants, and more difficult items. The positive association between response time and accuracy is consistent with a speed-accuracy trade-off in participants' performance. This interpretation is corroborated by the main effect of rest score, where better-performing participants were slower overall.

@@

In sum, we found evidence of a speed-accuracy trade-off in performance in this sample. Accurate responses and more accurate participants were both slower overall. Notably, better-performing participants demonstrated adjustments to their response times as function of item difficulty; whereas the lowest scoring participants maintained the same work rate regardless of item difficulty, the highest scoring participants showed the largest slowing in responding as items became more challenging. We return to these results in the General Discussion.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. By comparison, Model 5 demonstrated smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). This result suggests the presence of non-negligible residual variance in item discrimination across item templates but not clones. It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 5 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

@@

Contrary to the assumption that item clones are exchangeable, we found an association between distractor type and item difficulty ($\delta_{\beta_3}$ = 1.105, 95\% HDI = 0.940 -- 1.269). Items with MD distractors were associated with an estimated 19.9\% reduction in accuracy compared to their equivalent items with PD distractors. Together element number, rule number, and distractor type explained slightly over half (52.0\%) of the variance in item difficulty. Despite this item response models that included a clone-level residual variance term for item difficulty was preferred to models without. Thus, the results indicate that not all item clones are not equally difficult and are thus cannot be assumed to be exchangeable.

To better demonstrate the residual variance in item difficulty across clones, the estimated item difficulty per clone organized by distractor type and shape set is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast, the pattern of item difficulties across clones by shape set is more complicated (Figure \ref{fig:fig02}b). After controlling for distractor type and mean response time, the magnitude of the clone-level residual variance is $\epsilon_{\beta k}$ = 0.620 (95\% HDI = 0.538 -- 0.701). Across the 128 unique sets of clones (i.e. across item and distractor type), that translates into an average difference in difficulty of 0.752 (95\% HDI = 0.630 -- 0.877) between the easiest and hardest item clones of each set. In other words, the magnitude of residual variance in difficulty across clones is roughly two-thirds as large as the effect of distractor type. Importantly, the most difficult clone in each triad varied almost uniformly by shape set (shape set 1 = 39; shape set 2 = 32; shape set 3 = 57). Thus, the sources of residual variability in item difficulty across clones is complex and likely reflect the idiosyncratic contributions of low-level perceptual features.

Turning to item discrimination, we did not find a credible association between discrimination and distractor type ($\delta_{\alpha_3}$ = -0.029, 95\% HDI = -0.065 -- 0.005). Moreover, a model that included a clone-level random effects term for item discrimination was not preferred to a model without. As such, there is not sufficient evidence to reject the assumption that item clones are equivalently discriminating. We note this result should be interpreted with caution given the relatively weak predictive improvement of the best-fitting model compared to a model with a clone-level random effects term.

Finally, we inspected the relationship between mean response time and item functioning. As expected, there was a positive association between mean response time and item difficulty ($\delta_{\beta_4}$ = 0.541, 95\% HDI = 0.414 -- 0.671) indicating response times were slower for more difficult items. There were not a credible association between item discrimination and average response time ($\delta_{\alpha_4}$ = -0.010, 95\% HDI = -0.030 -- 0.009).

@@

In the best-fitting model, there were several credible associations between person-level attributes and ability. There was a negative association between age and ability ($\rho_1$ = -0.299, 95\% HDI = -0.353 -- -0.246). There was also a small association between gender and ability ($\rho_2$ = 0.128, 95\% HDI = 0.078 -- 0.182), such that women performed marginally better than men. Average response time was positively associated with ability ($\rho_3$ = 0.427, 95\% HDI = 0.374 -- 0.478), as was $\Delta$ response time ($\rho_4$ = 0.368, 95\% HDI = 0.313 -- 0.424). In other words participants with higher levels of ability spent more time on each item on average, and even longer for more difficult items. Together, these four person attributes explained 50.3\% of the variance in ability. These results are consistent with a speed-accuracy trade-off. That is, higher levels of ability in this sample reflect in part a tendency to slow down and respond carefully overall (and especially for the most challenging items).

@@

Finally, we validated the fit of the best-fitting model using two posterior predictive model checking measures. The PPP-value for chi-square discrepancy statistic did not exceed the critical threshold ($\chi^2_{NC}$ = 15.904, PPP-value = 0.297), indicating that the model was sufficiently able to reproduce the distribution of observed scores. The PPP-value for the SGDDM statistic did exceed the critical threshold (PPP-value = 0.019). However, the mean of the realized SGDDM values ($\text{SGDDM}_r$ = 0.0180) were only marginally larger than the mean of the posterior predicted SGDDM values ($\text{SGDDM}_p$ = 0.0175) meaning that the residual inter-item correlations in the data were small on average and only slightly larger than what we would expect to observe by chance. As such, there is little evidence for local dependence in the data. Overall then, we can conclude the best-fitting model provides an adequate fit to the data.

@@

In this first study, we recruited a large sample of adults to complete items from the MaRs-IB, making sure that each item template and clones were completed by a sizable number of participants. Using multilevel item structure models, we then estimated the psychometric properties (i.e. difficulty and discrimination) of each item. Corroborating the results of an initial investigation of the MaRs-IB \citep{chierchia2019matrix}, we found that the items in the MaRs-IB vary greatly in their difficulty -- a prerequisite for measuring nonverbal reasoning across the ability spectrum. We also found that the items in the MaRs-IB exhibited medium-to-large levels of discrimination, similar to what has been found in investigations of other of matrix reasoning tasks \citep{chiesi2012using, chiesi2012item, van2013shortened}.

We also investigated how item complexity impacts item functioning. We found that element number and rule number were both positively associated with item difficulty, the effects of which were of approximately equal magnitude (i.e. a one-unit change in either was independently associated with a roughly 10\% reduction in performance). This finding is interesting in light of previous investigations of nonverbal reasoning tasks, which have found a greater influence on item difficulty from either the number of elements \citep{bethell1984adaptive} or number of rules \citep{mulholland1980components}. One possible reason for this discrepancy is that these two attributes are largely uncorrelated ($\rho$ = 0.176) across items in the MaRs-IB, which allows for unconfounded estimates of their effects. Together element number and rule number explained 67.6\% of the variance in item difficulty across item templates (and 38.6\% of the variance across item clones), which is in line with previous investigations of matrix reasoning tasks \citep{carpenter1990one, matzen1994error}. These findings further validate the design of the MaRs-IB insofar that item complexity is a primary determinant of item difficulty.

Finally, we investigated whether the item clones in the MaRs-IB are psychometrically equivalent and exchangeable. Crucially, we found evidence that item clones are not equally difficult. Unexpectedly, and contrary to the results of \cite{chierchia2019matrix}, distractor type emerged as a substantial predictor of item difficulty (i.e. items with MD-type distractors are associated with a 19.9\% reduction in accuracy compared to the same items with PD-type distractors). This discrepancy in findings between studies may reflect our larger sample size (N=1501 versus N=659 participants) or population (adults-only versus a mixture of adults, adolescents, and children), and deserves further investigation. Even after controlling for element number, rule number, and distractor type, we found non-negligible residual variability in item difficulty across clones that was difficult to characterize and likely reflects both the additive and interactive effects of low-level perceptual features. Regardless, our results clearly indicate that item clones in the MaRs-IB cannot be assumed to be psychometrically equivalent and should not be treated as exchangeable.
